Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-03-05 main.Rmd
content cut short - clashes and harmonies

Firstly, empty combinations can arise due to the structure of the calendar or hierarchy.
These are "structurally" empty combinations in that it is impossible for them to have any value of the measured variable. Let us take a specific example, where $C_1$ be day-of-month with 31 levels and $C_2$ be week-of-month with 5 levels. There will be $31\times 5=155$ sets $S_{ij}$ corresponding to the possible combinations of day-of-month and week-of-month. Many of these are empty. For example $S_{1,5}$, $S_{21,2}$, etc. This is also intuitive since the first day of the month can never correspond to fifth week of the month.

@@

Thirdly, empty combinations can be a result of how granularities are constructed. These are called "build-based" empty combinations. Let $C_1$ be Business-days, which are days from Monday to Friday except holidays and $C_2$ is day-of-month. Then the days denoting weekends in a month would not correspond to any Business days. This is different from structurally empty combinations because structure of the calendar does not lead to these missing combinations, but the construction of the granularity does.

An example when there will be no empty combinations could be where $C_1$ and $C_2$ maps index set to day-of-week and month-of-year respectively. Here $C_1$ can have 7 levels while $C_2$ can have 12 levels. So there are $12\times7=84$ sets $S_{ij}$ corresponding to the possible combinations of day-of-week and month-of-year. All of these are non-empty because every day-of-week can occur in every month.

@@

# filter(fielding_wckts %in% c(0,1)) %>%
cricket_data %>% prob_plot("over", "lag_field",
  hierarchy_model,
  response = "run_rate",
  plot_type = "violin",
) + ggtitle("") + geom_boxplot(width = 0.5, aes(colour = lag_field)) + ylim(0,4.5) +
      theme(
      axis.text = element_text(size = 24),
      strip.text = element_text(size = 24),
      axis.title = element_text(size = 24),
      legend.title = element_text(size = 24),
      legend.text = element_text(size = 24),
      legend.position = "bottom",
      plot.title =  element_text(size = 24)
    )
```

@@

Q3: Are runs set to reduce in the next over for dot balls in the previous over?

A dot ball is a delivery bowled without any runs scored off it. The number of dot balls is reflective of the quality of bowling in the game. Run rate of an over should ideally decrease if the number of dot balls increase. However, what is the effect of dot balls on runs scored in the subsequent over. Will players  batsman likely to go for big shots because they couldn't score good runs in the previous over? Or they should play consistently and avoid scoring high? Figure \ref{fig:exdot} shows the quantile plot of runs across overs for at least one dot ball per over (facet 1) or no dot balls per over (facet 2).
With at least one dot balls per over, the distribution of run rates in facet 1 increase slower compared to that in facet 2. This implies that run rates are likely to decrease in the subsequent over as a result of dot balls in the previous over.

<!-- Each of the quantiles decrease across facets. With two dot balls per over, at least 50% of the times, run rates decrease in the subsequent over as observed through the negative values for different quantiles. -->

<!-- # ```{r crickread, ech0=FALSE} -->
<!-- # cricket_tsibble <- gravitas::cricket %>% -->
<!-- #   mutate(data_index = row_number()) %>% -->
<!-- #   as_tsibble(index = data_index) -->
<!-- # ``` -->

```{r exdot, fig.cap="25th, 50th, 75th quantiles of runs per over are drawn  across overs of the innings with no (facet 2), more than zero (facet 1) dot balls per over. For all quantiles, run rates mostly increase at a higher rate in facet 2 compared to facet 1 implying run rates decrease with at least one dot ball in the previous over.", fig.pos="ht"}

# library(gravitas)
# library(tsibble)
#
#
# cricket_bowling <- cricket_season %>%
#   mutate(dot_ball_proxy = if_else(total_runs==0,1,0),
#          wicket_proxy = if_else(is.na(dismissal_kind),0,1)) %>%
#   group_by(season, match_id, batting_team, bowling_team,  inning, over) %>%
#   summarise(runs_per_over = sum(total_runs),
#             run_rate = sum(total_runs)/length(total_runs),
#             dot_balls = sum(dot_ball_proxy)) %>%
#             mutate(diff_run_rate = c(0,diff(run_rate)))
#
#
# cricket_tsibble_bow <- cricket_bowling %>%
#   ungroup() %>%
#   mutate(data_index = row_number()) %>%
#   as_tsibble(index = data_index)
#

# cricket_tsibble_bow %>%
#   mutate(dot = if_else(dot_balls==0,"no dot balls", ">0 dot balls")) %>%
#   #filter(dot_balls %in% c(0, 1, 2)) %>%
#   prob_plot("dot",
#             "over",
#             hierarchy_model,
#             response = "diff_run_rate",
#             plot_type = "quantile",
#             quantile_prob = c(0.1, 0.2, 0.3, 0.4, 0.5),
#             symmetric = FALSE)  + ggtitle("")

cricket_data %>%
  # filter(dot_balls %in% c(0, 1, 2)) %>%
  prob_plot("lag_dot",
    "over",
    hierarchy_model,
    response = "run_rate",
    plot_type = "quantile",
    quantile_prob = c(0.25, 0.5, 0.75),
    symmetric = FALSE
  ) + ggtitle("")
```

In Q2 and Q3, dots balls per over or catches per over are considered as aperiodic cyclic granularities with dot balls or catches as aperiodic linear granularities. These aperiodic linear granularities do not appear in the hierarchy table since it is difficult to position them in a hierarchy. These are similar to holidays or special events in temporal data. While any special event that corresponds to a time domain can be treated as an aperiodic linear granularity in a temporal case, dot balls or wickets that corresponds to certain balls (index) could be treated as aperiodic events in cricket.

<!-- For example, in the temporal case, aperiodic cyclic granularity "Easter" is computed based on which all time indices correspond to the event "Easter". Similarly, for aperiodic cyclic granularity "dot balls per over" is computed based on which all balls correspond to the event "dot balls". -->

@@

>>>>

Firstly, empty combinations can arise due to the structure of the calendar or hierarchy. These are called "structurally" empty combinations. Let us take a specific example, where $C_1$ be day-of-month with 31 levels and $C_2$ be week-of-month with 5 levels. There will be $31\times 5=155$ sets $S_{ij}$ corresponding to possible combinations of $C_1$ and $C_2$. Many of these like $S_{1,5}$, $S_{21,2}$ are empty. This is also intuitive since the first day of the month can never correspond to fifth week of the month.

@@

Thirdly, empty combinations can be a result of how granularities are constructed. These are called "build-based" empty combinations. Let $C_1$ be Business-days, which are days from Monday to Friday except holidays and $C_2$ be day-of-month. Then the days denoting weekends in a month would not correspond to any Business days. This is different from structurally empty combinations because structure of the calendar does not lead to these missing combinations, but the construction of the granularity does.

An example when there will be no empty combinations could be where $C_1$ and $C_2$ maps index set to day-of-week and month-of-year respectively. Here $C_1$ can have 7 levels while $C_2$ can have 12 levels. So there are $12\times7=84$ sets $S_{ij}$. All of these are non-empty because every day-of-week can occur in every month.

@@

# filter(fielding_wckts %in% c(0,1)) %>%
cricket_data %>% prob_plot("over", "lag_field",
  hierarchy_model,
  response = "run_rate",
  plot_type = "violin",
) + ggtitle("") + geom_boxplot(width = 0.5, aes(colour = lag_field)) + ylim(0,4.5)
```

@@

<!-- Q3: Are runs set to reduce in the next over for dot balls in the previous over? -->

<!-- A dot ball is a delivery bowled without any runs scored off it. The number of dot balls is reflective of the quality of bowling in the game. Run rate of an over should ideally decrease if the number of dot balls increase. However, what is the effect of dot balls on runs scored in the subsequent over. Will players  batsman likely to go for big shots because they couldn't score good runs in the previous over? Or they should play consistently and avoid scoring high? Figure \ref{fig:exdot} shows the quantile plot of runs across overs for at least one dot ball per over (facet 1) or no dot balls per over (facet 2). -->
<!-- With at least one dot balls per over, the distribution of run rates in facet 1 increase slower compared to that in facet 2. This implies that run rates are likely to decrease in the subsequent over as a result of dot balls in the previous over. -->

<!-- ```{r exdot, fig.cap="25th, 50th, 75th quantiles of runs per over are drawn  across overs of the innings with no (facet 2), more than zero (facet 1) dot balls per over. For all quantiles, run rates mostly increase at a higher rate in facet 2 compared to facet 1 implying run rates decrease with at least one dot ball in the previous over.", fig.pos="ht"} -->

<!-- cricket_data %>% -->
<!--   # filter(dot_balls %in% c(0, 1, 2)) %>% -->
<!--   prob_plot("lag_dot", -->
<!--     "over", -->
<!--     hierarchy_model, -->
<!--     response = "run_rate", -->
<!--     plot_type = "quantile", -->
<!--     quantile_prob = c(0.25, 0.5, 0.75), -->
<!--     symmetric = FALSE -->
<!--   ) + ggtitle("") -->
<!-- ``` -->

In Q2, wickets per over are considered as aperiodic cyclic granularities with dot balls or catches as aperiodic linear granularities. These aperiodic linear granularities do not appear in the hierarchy table since it is difficult to position them in a hierarchy. These are similar to holidays or special events in temporal data. While any special event that corresponds to a time domain can be treated as an aperiodic linear granularity in a temporal case, dot balls or wickets that corresponds to certain balls (index) could be treated as aperiodic events in cricket.

@@

# Appendix
